---
title: "Introduction to ctdR"
author:
  - name: Luigi Corsaro
    affiliation: ""
    orcid: 0000-0003-1218-230X
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to ctdR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Overview

**ctdR** identifies chemicals significantly associated
with a set of genes using data from the
[Comparative Toxicogenomics Database (CTD)](https://ctdbase.org).

Four enrichment methods are supported through a single
`enrichment_CTD()` interface; the input shape depends on
the method:

- **ORA** (Over-Representation Analysis) -- gene list +
    hypergeometric test. Powered by
    `clusterProfiler::enricher`.
- **GSEA** (Gene Set Enrichment Analysis) -- ranked gene
    list + permutation test. Powered by `fgsea::fgsea`.
- **CAMERA** (competitive gene-set test) -- expression
    matrix + design + contrast. Corrects for inter-gene
    correlation within each chemical's gene set. Powered
    by `limma::camera`.
- **GSVA** (Gene Set Variation Analysis) -- expression
    matrix + per-sample scoring. Returns a chemical x sample
    score matrix. Powered by `GSVA::gsva`.

# Data licensing disclaimer

> **This package does NOT bundle, redistribute, or embed
> any data from the Comparative Toxicogenomics Database.**
> CTD data are created and maintained by NC State
> University and are subject to specific licensing terms.
> Users must download the data directly from
> <https://ctdbase.org> and comply with the
> [CTD Terms of Service](https://ctdbase.org/about/legal.jsp).

# Installation

> 🚧 ctdR is *not yet accepted, or under review* for Bioconductor
> (submission [#4232](https://github.com/Bioconductor/Contributions/issues/4232)).
> Until then, please install from GitHub. *Fingers crossed* 🤞.

## From GitHub (current installation method)

    # install.packages("devtools")
    devtools::install_github("drake69/ctdR")

## From Bioconductor (once accepted)

Once ctdR is accepted into Bioconductor (currently under review), it
will be installable directly via:

    BiocManager::install("ctdR")

# Quick start

The ctdR workflow has three steps:

1. **Download** the CTD data file (once, manually)
2. **Import** the data into ctdR (once)
3. **Analyse** your gene list (as many times as needed)

## Step 1 -- Download CTD data

Download **`CTD_chem_gene_ixns.csv.gz`** from
<https://ctdbase.org/reports/CTD_chem_gene_ixns.csv.gz>.

Then decompress it:

```bash
gunzip CTD_chem_gene_ixns.csv.gz
```

This produces `CTD_chem_gene_ixns.csv` (several GB
uncompressed).

## Step 2 -- Import into ctdR

In production you would run:

```{r import-real, eval=FALSE}
library(ctdR)
import_CTD("~/Downloads/CTD_chem_gene_ixns.csv")
```

For this vignette we use a small **synthetic** dataset
bundled with the package:

```{r import-sample}
library(ctdR)
sample_file <- system.file(
    "extdata", "CTD_chem_gene_ixns_sample.csv",
    package = "ctdR"
)
import_CTD(sample_file)
```

`import_CTD()` performs the following:

1. Reads the CSV (skipping the 27 CTD header lines).
2. Filters interactions to **Homo sapiens** only
    (OrganismID 9606).
3. Collects Entrez gene IDs for each chemical.
4. Maps Entrez IDs to HGNC gene symbols via
    `org.Hs.eg.db`.
5. Caches the processed data locally
    (under `rappdirs::user_cache_dir("ctdR")`).

This step takes several minutes with the full CTD file.
You only need to run it once -- or again when you
download a newer CTD release.

## Step 3 -- Run enrichment analysis

### Prepare your gene list

Your input must be a data frame with at least two
columns:

| Column | Description |
|---|---|
| `EntrezID` | Character or numeric Entrez gene IDs |
| *(2nd column)* | Numeric value per gene (e.g. p-value) |

The second column is used for ranking in GSEA and is
ignored in ORA.

```{r prepare-genes}
genes <- data.frame(
    EntrezID = c(
        "7124", "3569", "7157", "672", "1956",
        "4609", "3845", "207", "5290", "3553"
    ),
    pvalue = c(
        0.001, 0.003, 0.005, 0.008, 0.01,
        0.02, 0.03, 0.04, 0.05, 0.06
    )
)
```

### Shared output schema

All three data-frame-returning methods (ORA, GSEA,
CAMERA) share the **same leading columns**. This makes
it trivial to combine results across methods (e.g.
`dplyr::bind_rows()` on a list of result frames).

| Column | Type | Description |
|---|---|---|
| `ChemicalID` | character | CTD chemical identifier |
| `ChemicalName` | character | Human-readable chemical name |
| `Method` | character | One of `"ORA"`, `"GSEA"`, `"CAMERA"` |
| `PValue` | numeric | Raw p-value from the underlying method |
| `PValueAdjusted` | numeric | Multiple-testing-corrected p-value (`pAdjustMethod`) |

Method-specific extras follow these front columns and
differ by method (documented in the sub-sections
below). Rows are sorted by `PValueAdjusted` ascending.

### Over-Representation Analysis (ORA)

ORA tests whether the overlap between your gene list
and each chemical's known gene targets is significantly
larger than expected by chance.

```{r ora}
ora_results <- enrichment_CTD(genes, method = "ORA")
head(ora_results)
```

**ORA method-specific columns** (in addition to the
shared schema above):

| Column | Description |
|---|---|
| `GeneRatio` | Proportion of input genes in the set |
| `BackgroundRatio` | Background ratio |
| `QValue` | Storey's q-value (from `clusterProfiler`) |
| `FoldEnrichment` | GeneRatio / BackgroundRatio |
| `EnrichedGenes` | Enriched gene symbols (comma-separated) |
| `Count` | Number of overlapping genes |

### Gene Set Enrichment Analysis (GSEA)

GSEA uses the full ranked gene list to detect chemicals
whose targets cluster toward the top or bottom of the
ranking. For best results, supply a signed ranking
statistic (e.g. the moderated t-statistic from
`limma::eBayes()`) as a column named `stat`. When `stat`
is absent, ctdR falls back to `-log10(pvalue)`, which
loses directionality and may produce ties.

```{r gsea}
# Quick-start: no signed statistic available in this
# toy gene list, so the pvalue fallback is used.
gsea_results <- enrichment_CTD(genes, method = "GSEA")
head(gsea_results)
```

**GSEA method-specific columns** (in addition to the
shared schema above):

| Column | Description |
|---|---|
| `EnrichmentScore` | Enrichment score (`ES` from fgsea) |
| `NormalizedEnrichmentScore` | Normalized enrichment score (`NES`) |
| `GeneSetSize` | Size of the gene set used |
| `LeadingEdge` | Leading-edge gene subset |
| `FoldEnrichment` | `abs(ES) / mean(ES)` |
| `EnrichedGenes` | Comma-separated enriched genes |

### End-to-end example with real RNA-seq data (GSE311566)

The hand-coded `genes` data frame above is convenient for the
Quick start but does not exercise the matrix-based methods. This
section ties everything together on a small subset of
[GEO GSE311566](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE311566)
-- human PBMCs treated with **dexamethasone** vs **vehicle (DMSO)**
in female donors. The bundled subset
`inst/extdata/GSE311566_subset.rds` contains 1,500 top-variance
genes (plus the 17 genes referenced by the toy CTD sample),
log2-transformed normalised counts, 7 samples total
(4 DMSO + 3 Dex). See `inst/extdata/README.md` for provenance.

```{r load-real}
gse <- readRDS(system.file(
    "extdata", "GSE311566_subset.rds", package = "ctdR"
))
expr <- gse$expr
grp  <- gse$coldata$group
dim(expr)
table(grp)
```

We compute a deliberately minimal per-gene differential
expression with base R only: a two-sample `t.test` per gene with
BH-adjusted p-values. Production analyses should use `limma`,
`DESeq2`, or `edgeR` for proper count-based modelling; this
ascetic version keeps the example self-contained.

```{r de-real}
de <- t(apply(expr, 1, function(y) {
    tt <- stats::t.test(y ~ grp)
    c(log2FC = unname(diff(tt$estimate)),
      pvalue = tt$p.value)
}))
de <- as.data.frame(de)
de$padj <- stats::p.adjust(de$pvalue, method = "BH")
de$EntrezID <- rownames(de)
de <- de[, c("EntrezID", "log2FC", "pvalue", "padj")]
head(de[order(de$padj), ])
c(padj_lt_05 = sum(de$padj < 0.05),
  pvalue_lt_05 = sum(de$pvalue < 0.05))
```

#### Real-data GSEA

GSEA uses the full ranked gene list and tends to surface the
expected chemical (Dexamethasone) near the top even with this
small DE. We supply `log2FC` as the `stat` column so that
direction (up/down) is preserved in the ranking.

```{r gsea-real}
gsea_real <- enrichment_CTD(
    data.frame(EntrezID = de$EntrezID,
               pvalue   = de$pvalue,
               stat     = de$log2FC),
    method = "GSEA"
)
head(gsea_real[, c("ChemicalID", "ChemicalName",
                   "PValue", "PValueAdjusted",
                   "NormalizedEnrichmentScore")])
```

#### Real-data ORA

ORA is statistically meaningful only with the **full CTD download**
(~16,000 chemicals × millions of gene interactions, available at
<https://ctdbase.org/reports/CTD_chem_gene_ixns.csv.gz>). The bundled
`CTD_chem_gene_ixns_sample.csv` covers 10 chemicals × 17 genes — its
sole purpose is API demonstration. With only 7 samples and a universe
this small, ORA is expected to return zero enriched chemicals. Set
`is_subset <- FALSE` in the chunk below to run a fully powered analysis
on the real CTD. All other methods (GSEA, CAMERA, GSVA) work correctly
with the bundled data.

```{r ora-is-subset}
## Set is_subset <- FALSE to run ORA on the full CTD download.
## Requires CTD_chem_gene_ixns.csv.gz from ctdbase.org, decompressed locally.
is_subset <- TRUE
```

```{r ora-real}
sig <- de[de$pvalue < 0.05, c("EntrezID", "pvalue")]

if (is_subset) {
    ora_real <- enrichment_CTD(sig, method = "ORA")
    if (nrow(ora_real)) {
        head(ora_real[, c("ChemicalID", "ChemicalName",
                          "PValue", "PValueAdjusted", "Count")])
    } else {
        message(
            "ORA returned 0 results on the toy CTD sample (expected).\n",
            "Set is_subset <- FALSE and provide CTD_chem_gene_ixns.csv",
            " for a powered analysis."
        )
    }
} else {
    ## Full CTD path -----------------------------------------------------------
    ## 1. Download https://ctdbase.org/reports/CTD_chem_gene_ixns.csv.gz
    ## 2. gunzip CTD_chem_gene_ixns.csv.gz
    ## 3. Set ctd_path to the decompressed file location
    ctd_path <- "~/Downloads/CTD_chem_gene_ixns.csv"  # adjust to your path
    import_CTD(ctd_path)
    ora_full <- enrichment_CTD(sig, method = "ORA")
    if (nrow(ora_full)) {
        head(ora_full[order(ora_full$PValue),
                      c("ChemicalID", "ChemicalName",
                        "PValue", "PValueAdjusted",
                        "FoldEnrichment", "Count")])
    } else {
        message("No chemicals enriched — check CTD file path and gene list.")
    }
}
```

### CAMERA (competitive test with inter-gene correlation)

CAMERA tests, for each chemical, whether its target
genes show a stronger differential signal than the rest
of the transcriptome under a user-supplied design and
contrast, while correcting for inter-gene correlation
within the gene set. Reusing the real `expr` / `grp` above:

```{r camera}
design <- model.matrix(~ grp)

camera_results <- enrichment_CTD(
    expr,
    method   = "CAMERA",
    design   = design,
    contrast = 2  # column 2 of design = Dex vs DMSO
)
head(camera_results)
```

**CAMERA method-specific columns** (in addition to the
shared schema above):

| Column | Description |
|---|---|
| `GeneSetSize` | Gene set size after intersection with `rownames(x)` |
| `Direction` | `"Up"` or `"Down"` |
| `Correlation` | Inter-gene correlation (present only when estimated by `limma::camera`; absent when `inter.gene.cor` is fixed) |

### GSVA (per-sample scoring)

GSVA produces a **per-sample** enrichment score for each
chemical, returning a matrix (chemicals in rows, samples
in columns) suitable for clustering, association tests
against phenotypes, survival analysis, or heatmap
visualization.

```{r gsva}
gsva_scores <- enrichment_CTD(expr, method = "GSVA")
dim(gsva_scores)  # chemicals x samples
head(gsva_scores, 3)
```

Tune the underlying `GSVA::gsvaParam()` through `...`,
for example to restrict to gene sets of a given size:

```{r gsva-tune}
gsva_strict <- enrichment_CTD(
    expr, method = "GSVA",
    minSize = 5, maxSize = 500
)
nrow(gsva_strict)
```

## Visualizing results

`plot_CTD()` creates publication-ready plots from enrichment
results. It auto-detects the method: bar/dot plots of fold
enrichment for ORA/GSEA, bar/dot plots of `-log10(padj)`
coloured by direction for CAMERA, and a sample-level
heatmap of the top-variance chemicals for GSVA.

```{r plot-bar, fig.width=7, fig.height=5}
# ORA bar plot of top enriched chemicals
plot_CTD(ora_results, type = "bar")
```

```{r plot-dot, fig.width=7, fig.height=5}
# ORA dot plot (size = gene count, color = adjusted p-value)
plot_CTD(ora_results, type = "dot", n = 10)
```

```{r plot-camera, fig.width=7, fig.height=5}
# CAMERA bar plot: x-axis = -log10(padj), fill = Direction
plot_CTD(camera_results, type = "bar")
```

```{r plot-gsva, fig.width=7, fig.height=5}
# GSVA heatmap: top-variance chemicals across samples
plot_CTD(gsva_scores)
```

## Adjusting for multiple testing

By default, `enrichment_CTD()` uses the Benjamini-Hochberg
(`"BH"`) method for p-value adjustment. You can change this
via the `pAdjustMethod` parameter:

```{r padj-methods}
# Bonferroni correction (more conservative)
ora_bonf <- enrichment_CTD(genes, method = "ORA",
    pAdjustMethod = "bonferroni")

# No adjustment
ora_raw <- enrichment_CTD(genes, method = "ORA",
    pAdjustMethod = "none")

# Compare the adjusted p-value of the top hit
data.frame(
    method = c("BH (default)", "bonferroni", "none"),
    top_padj = c(min(ora_results$padj),
        min(ora_bonf$padj), min(ora_raw$padj))
)
```

Available methods: `"BH"` (default), `"bonferroni"`,
`"fdr"` (alias for BH), `"none"`.

## Gene set size filters and background universe

Each method applies a gene set size filter and defines a background
universe, but the degree of user control differs:

| Method | Size filter | Configurable | Background universe | Configurable |
|--------|-------------|:---:|---------------------|:---:|
| ORA | `minGSSize` / `maxGSSize` | ✓ | All genes in TERM2GENE (default) or user-supplied `universe` | ✓ |
| GSEA | `minSize` / `maxSize` | ✓ | No universe concept — uses the full ranked list | — |
| CAMERA | ≥ 2 genes (hard-coded) | ✗ | `rownames(expr)` | implicit |
| GSVA | `minSize` / `maxSize` | ✓ | `rownames(expr)` | implicit |

ORA, GSEA, and GSVA accept size-filter arguments via `...`, which are
forwarded to the underlying engine. CAMERA's minimum of 2 genes is
hard-coded and cannot be changed.

```{r size-universe}
# ORA: restrict background to expressed genes and apply size filters
ora_expressed <- enrichment_CTD(
    genes, method = "ORA",
    universe   = c(genes$EntrezID, "7422", "836"),  # expressed genes
    minGSSize  = 3,
    maxGSSize  = 300
)

# GSEA: filter gene sets by size
gsea_filtered <- enrichment_CTD(
    genes, method = "GSEA",
    minSize = 3,
    maxSize = 300
)
```

Setting `universe` in ORA to the full list of genes measured in your
experiment (rather than the default of all CTD-annotated genes) is
strongly recommended: it avoids inflated fold-enrichment estimates that
arise when the background is larger than the actual measurement space.

# Choosing among the four methods

| Aspect | ORA | GSEA | CAMERA | GSVA |
|---|---|---|---|---|
| **Input** | Gene list | Ranked gene list | Expression matrix + design + contrast | Expression matrix |
| **Question** | Over-represented? | Cluster at extremes? | Stronger signal than rest of transcriptome (correlation-corrected)? | Per-sample chemical score |
| **Output** | Ranked chemicals (data.frame) | Ranked chemicals (data.frame) | Ranked chemicals (data.frame) | Chemical x sample score matrix |
| **Best for** | DEG lists | Exploratory, ranked | Multi-sample DE experiments, co-regulated gene sets | Patient stratification, downstream tests, heatmaps |

Use **ORA** when you have a well-defined gene list above a
significance threshold; **GSEA** to leverage the full
ranking without a cutoff; **CAMERA** when you have a
proper multi-sample design and want to control the
inflated false-positive rate that ORA/GSEA exhibit on
strongly co-regulated gene sets; **GSVA** when you want a
per-sample score to feed into downstream analyses
(clustering, survival, correlation with phenotypes).

# Updating the CTD data

CTD releases updated data periodically. To update:

1. Download the latest `CTD_chem_gene_ixns.csv.gz` from
    <https://ctdbase.org/reports/CTD_chem_gene_ixns.csv.gz>.
2. Decompress and re-run `import_CTD()` -- existing
    cache files are overwritten.

```{r reimport, eval=FALSE}
import_CTD("~/Downloads/CTD_chem_gene_ixns.csv")
```

# Session info

```{r session-info}
sessionInfo()
```
