---
title: "Introduction to ctdR"
author:
  - name: Luigi Corsaro
    affiliation: ""
    orcid: 0000-0003-1218-230X
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to ctdR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Overview

**ctdR** identifies chemicals significantly associated
with a set of genes using data from the
[Comparative Toxicogenomics Database (CTD)](https://ctdbase.org).

Two enrichment methods are supported:

- **ORA** (Over-Representation Analysis) -- tests whether
    the overlap between your gene list and each chemical's
    target genes is larger than expected by chance.
    Powered by `clusterProfiler::enricher`.
- **GSEA** (Gene Set Enrichment Analysis) -- detects
    chemicals whose targets cluster toward the extremes
    of a ranked gene list. Powered by `fgsea::fgsea`.

# Data licensing disclaimer

> **This package does NOT bundle, redistribute, or embed
> any data from the Comparative Toxicogenomics Database.**
> CTD data are created and maintained by NC State
> University and are subject to specific licensing terms.
> Users must download the data directly from
> <https://ctdbase.org> and comply with the
> [CTD Terms of Service](https://ctdbase.org/about/legal.jsp).

# Installation

## From Bioconductor

    BiocManager::install("ctdR")

## From GitHub (development version)

    # install.packages("devtools")
    devtools::install_github("drake69/ctdR")

# Quick start

The ctdR workflow has three steps:

1. **Download** the CTD data file (once, manually)
2. **Import** the data into ctdR (once)
3. **Analyse** your gene list (as many times as needed)

## Step 1 -- Download CTD data

Download **`CTD_chem_gene_ixns.csv.gz`** from
<https://ctdbase.org/reports/CTD_chem_gene_ixns.csv.gz>.

Then decompress it:

```bash
gunzip CTD_chem_gene_ixns.csv.gz
```

This produces `CTD_chem_gene_ixns.csv` (several GB
uncompressed).

## Step 2 -- Import into ctdR

In production you would run:

```{r import-real, eval=FALSE}
library(ctdR)
import_CTD("~/Downloads/CTD_chem_gene_ixns.csv")
```

For this vignette we use a small **synthetic** dataset
bundled with the package:

```{r import-sample}
library(ctdR)
sample_file <- system.file(
    "extdata", "CTD_chem_gene_ixns_sample.csv",
    package = "ctdR"
)
import_CTD(sample_file)
```

`import_CTD()` performs the following:

1. Reads the CSV (skipping the 27 CTD header lines).
2. Filters interactions to **Homo sapiens** only
    (OrganismID 9606).
3. Collects Entrez gene IDs for each chemical.
4. Maps Entrez IDs to HGNC gene symbols via
    `org.Hs.eg.db`.
5. Caches the processed data locally
    (under `rappdirs::user_cache_dir("ctdR")`).

This step takes several minutes with the full CTD file.
You only need to run it once -- or again when you
download a newer CTD release.

## Step 3 -- Run enrichment analysis

### Prepare your gene list

Your input must be a data frame with at least two
columns:

| Column | Description |
|---|---|
| `entrez_ids` | Character or numeric Entrez gene IDs |
| *(2nd column)* | Numeric value per gene (e.g. p-value) |

The second column is used for ranking in GSEA and is
ignored in ORA.

```{r prepare-genes}
genes <- data.frame(
    entrez_ids = c(
        "7124", "3569", "7157", "672", "1956",
        "4609", "3845", "207", "5290", "3553"
    ),
    pvalue = c(
        0.001, 0.003, 0.005, 0.008, 0.01,
        0.02, 0.03, 0.04, 0.05, 0.06
    )
)
```

### Over-Representation Analysis (ORA)

ORA tests whether the overlap between your gene list
and each chemical's known gene targets is significantly
larger than expected by chance.

```{r ora}
ora_results <- enrichment_CTD(genes, method = "ORA")
head(ora_results)
```

**ORA output columns:**

| Column | Description |
|---|---|
| `ChemicalID` | CTD chemical identifier |
| `ChemicalName` | Human-readable chemical name |
| `GeneRatio` | Proportion of input genes in the set |
| `BgRatio` | Background ratio |
| `pvalue` | Raw p-value |
| `padj` | Adjusted p-value (BH method) |
| `foldEnrichment` | GeneRatio / BgRatio |
| `geneID` | Enriched gene symbols |
| `Count` | Number of overlapping genes |

### Gene Set Enrichment Analysis (GSEA)

GSEA uses the full ranked gene list (ranked by the
numeric column) to detect chemicals whose targets
cluster toward the top or bottom of the ranking.

```{r gsea}
gsea_results <- enrichment_CTD(genes, method = "GSEA")
head(gsea_results)
```

**GSEA output columns:**

| Column | Description |
|---|---|
| `ChemicalID` | CTD chemical identifier |
| `ChemicalName` | Human-readable chemical name |
| `pval` | Raw p-value |
| `padj` | Adjusted p-value |
| `ES` | Enrichment score |
| `NES` | Normalized enrichment score |
| `size` | Size of the gene set |
| `leadingEdge` | Leading-edge gene subset |
| `foldEnrichment` | abs(ES) / mean(ES) |
| `Enriched_GENE` | Comma-separated enriched genes |

## Visualizing results

`plot_CTD()` creates publication-ready plots from enrichment
results. It automatically detects whether the input comes
from ORA or GSEA.

```{r plot-bar, fig.width=7, fig.height=5}
# Bar plot of top enriched chemicals
plot_CTD(ora_results, type = "bar")
```

```{r plot-dot, fig.width=7, fig.height=5}
# Dot plot (size = gene count, color = adjusted p-value)
plot_CTD(ora_results, type = "dot", n = 10)
```

## Adjusting for multiple testing

By default, `enrichment_CTD()` uses the Benjamini-Hochberg
(`"BH"`) method for p-value adjustment. You can change this
via the `pAdjustMethod` parameter:

```{r padj-methods, eval=FALSE}
# Bonferroni correction (more conservative)
enrichment_CTD(genes, method = "ORA", pAdjustMethod = "bonferroni")

# No adjustment
enrichment_CTD(genes, method = "ORA", pAdjustMethod = "none")
```

Available methods: `"BH"` (default), `"bonferroni"`,
`"fdr"` (alias for BH), `"none"`.

# Choosing between ORA and GSEA

| Aspect | ORA | GSEA |
|---|---|---|
| **Input** | Gene list (no ranking) | Ranked gene list |
| **Question** | Over-represented? | Cluster at extremes? |
| **Threshold** | Requires cutoff | No cutoff needed |
| **Best for** | DEG lists | Exploratory, ranked |

Use **ORA** when you have a well-defined gene list
(e.g. differentially expressed genes above a significance
threshold). Use **GSEA** when you want to leverage the
full ranking without imposing an arbitrary threshold.

# Updating the CTD data

CTD releases updated data periodically. To update:

1. Download the latest `CTD_chem_gene_ixns.csv.gz` from
    <https://ctdbase.org/reports/CTD_chem_gene_ixns.csv.gz>.
2. Decompress and re-run `import_CTD()` -- existing
    cache files are overwritten.

```{r reimport, eval=FALSE}
import_CTD("~/Downloads/CTD_chem_gene_ixns.csv")
```

# Session info

```{r session-info}
sessionInfo()
```
