---
title: "Introduction to ctdR"
author:
  - name: Luigi Corsaro
    affiliation: ""
    orcid: 0000-0003-1218-230X
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to ctdR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Overview

**ctdR** identifies chemicals significantly associated
with a set of genes using data from the
[Comparative Toxicogenomics Database (CTD)](https://ctdbase.org).

Four enrichment methods are supported through a single
`enrichment_CTD()` interface; the input shape depends on
the method:

- **ORA** (Over-Representation Analysis) -- gene list +
    hypergeometric test. Powered by
    `clusterProfiler::enricher`.
- **GSEA** (Gene Set Enrichment Analysis) -- ranked gene
    list + permutation test. Powered by `fgsea::fgsea`.
- **CAMERA** (competitive gene-set test) -- expression
    matrix + design + contrast. Corrects for inter-gene
    correlation within each chemical's gene set. Powered
    by `limma::camera`.
- **GSVA** (Gene Set Variation Analysis) -- expression
    matrix + per-sample scoring. Returns a chemical x sample
    score matrix. Powered by `GSVA::gsva`.

# Data licensing disclaimer

> **This package does NOT bundle, redistribute, or embed
> any data from the Comparative Toxicogenomics Database.**
> CTD data are created and maintained by NC State
> University and are subject to specific licensing terms.
> Users must download the data directly from
> <https://ctdbase.org> and comply with the
> [CTD Terms of Service](https://ctdbase.org/about/legal.jsp).

# Installation

> 🚧 ctdR is *not yet accepted, or under review* for Bioconductor
> (submission [#4232](https://github.com/Bioconductor/Contributions/issues/4232)).
> Until then, please install from GitHub. *Fingers crossed* 🤞.

## From GitHub (current installation method)

    # install.packages("devtools")
    devtools::install_github("drake69/ctdR")

## From Bioconductor (once accepted)

Once ctdR is accepted into Bioconductor (currently under review), it
will be installable directly via:

    BiocManager::install("ctdR")

# Quick start

The ctdR workflow has three steps:

1. **Download** the CTD data file (once, manually)
2. **Import** the data into ctdR (once)
3. **Analyse** your gene list (as many times as needed)

## Step 1 -- Download CTD data

Download **`CTD_chem_gene_ixns.csv.gz`** from
<https://ctdbase.org/reports/CTD_chem_gene_ixns.csv.gz>.

Then decompress it:

```bash
gunzip CTD_chem_gene_ixns.csv.gz
```

This produces `CTD_chem_gene_ixns.csv` (several GB
uncompressed).

## Step 2 -- Import into ctdR

In production you would run:

```{r import-real, eval=FALSE}
library(ctdR)
import_CTD("~/Downloads/CTD_chem_gene_ixns.csv")
```

For this vignette we use a small **synthetic** dataset
bundled with the package:

```{r import-sample}
library(ctdR)
sample_file <- system.file(
    "extdata", "CTD_chem_gene_ixns_sample.csv",
    package = "ctdR"
)
import_CTD(sample_file)
```

`import_CTD()` performs the following:

1. Reads the CSV (skipping the 27 CTD header lines).
2. Filters interactions to **Homo sapiens** only
    (OrganismID 9606).
3. Collects Entrez gene IDs for each chemical.
4. Maps Entrez IDs to HGNC gene symbols via
    `org.Hs.eg.db`.
5. Caches the processed data locally
    (under `rappdirs::user_cache_dir("ctdR")`).

This step takes several minutes with the full CTD file.
You only need to run it once -- or again when you
download a newer CTD release.

## Step 3 -- Run enrichment analysis

### Prepare your gene list

Your input must be a data frame with at least two
columns:

| Column | Description |
|---|---|
| `EntrezID` | Character or numeric Entrez gene IDs |
| *(2nd column)* | Numeric value per gene (e.g. p-value) |

The second column is used for ranking in GSEA and is
ignored in ORA.

```{r prepare-genes}
genes <- data.frame(
    EntrezID = c(
        "7124", "3569", "7157", "672", "1956",
        "4609", "3845", "207", "5290", "3553"
    ),
    pvalue = c(
        0.001, 0.003, 0.005, 0.008, 0.01,
        0.02, 0.03, 0.04, 0.05, 0.06
    )
)
```

### Shared output schema

All three data-frame-returning methods (ORA, GSEA,
CAMERA) share the **same leading columns**. This makes
it trivial to combine results across methods (e.g.
`dplyr::bind_rows()` on a list of result frames).

| Column | Type | Description |
|---|---|---|
| `ChemicalID` | character | CTD chemical identifier |
| `ChemicalName` | character | Human-readable chemical name |
| `Method` | character | One of `"ORA"`, `"GSEA"`, `"CAMERA"` |
| `PValue` | numeric | Raw p-value from the underlying method |
| `PValueAdjusted` | numeric | Multiple-testing-corrected p-value (`pAdjustMethod`) |

Method-specific extras follow these front columns and
differ by method (documented in the sub-sections
below). Rows are sorted by `PValueAdjusted` ascending.

### Over-Representation Analysis (ORA)

ORA tests whether the overlap between your gene list
and each chemical's known gene targets is significantly
larger than expected by chance.

```{r ora}
ora_results <- enrichment_CTD(genes, method = "ORA")
head(ora_results)
```

**ORA method-specific columns** (in addition to the
shared schema above):

| Column | Description |
|---|---|
| `GeneRatio` | Proportion of input genes in the set |
| `BackgroundRatio` | Background ratio |
| `QValue` | Storey's q-value (from `clusterProfiler`) |
| `FoldEnrichment` | GeneRatio / BackgroundRatio |
| `EnrichedGenes` | Enriched gene symbols (comma-separated) |
| `Count` | Number of overlapping genes |

### Gene Set Enrichment Analysis (GSEA)

GSEA uses the full ranked gene list (ranked by the
numeric column) to detect chemicals whose targets
cluster toward the top or bottom of the ranking.

```{r gsea}
gsea_results <- enrichment_CTD(genes, method = "GSEA")
head(gsea_results)
```

**GSEA method-specific columns** (in addition to the
shared schema above):

| Column | Description |
|---|---|
| `EnrichmentScore` | Enrichment score (`ES` from fgsea) |
| `NormalizedEnrichmentScore` | Normalized enrichment score (`NES`) |
| `GeneSetSize` | Size of the gene set used |
| `LeadingEdge` | Leading-edge gene subset |
| `FoldEnrichment` | `abs(ES) / mean(ES)` |
| `EnrichedGenes` | Comma-separated enriched genes |

### Build a synthetic expression matrix for CAMERA and GSVA

CAMERA and GSVA take a gene x sample expression matrix
(typically log2 counts or microarray intensities). For
this vignette we build a small synthetic matrix using the
Entrez IDs already loaded by `import_CTD()` above:

```{r make-expr}
# Pull the Entrez IDs covered by the bundled sample
cache_dir <- rappdirs::user_cache_dir("ctdR")
e <- new.env(parent = emptyenv())
load(file.path(cache_dir, "ChemicalName_GeneEntrezIds.rda"),
    envir = e)
all_ids <- as.character(unique(unlist(
    e$ChemicalName_GeneEntrezIds)))

# Synthetic expression matrix: 17 genes x 6 samples
set.seed(2026)
expr <- matrix(
    rnorm(length(all_ids) * 6),
    nrow = length(all_ids), ncol = 6,
    dimnames = list(all_ids, paste0("S", 1:6))
)
dim(expr)
```

### CAMERA (competitive test with inter-gene correlation)

CAMERA tests, for each chemical, whether its target
genes show a stronger differential signal than the rest
of the transcriptome under a user-supplied design and
contrast, while correcting for inter-gene correlation
within the gene set.

```{r camera}
grp <- factor(rep(c("ctrl", "treat"), each = 3))
design <- model.matrix(~ grp)

camera_results <- enrichment_CTD(
    expr,
    method   = "CAMERA",
    design   = design,
    contrast = 2  # last column of design = treat vs ctrl
)
head(camera_results)
```

**CAMERA method-specific columns** (in addition to the
shared schema above):

| Column | Description |
|---|---|
| `GeneSetSize` | Gene set size after intersection with `rownames(x)` |
| `Direction` | `"Up"` or `"Down"` |
| `Correlation` | Inter-gene correlation (when reported by `limma::camera`) |

### GSVA (per-sample scoring)

GSVA produces a **per-sample** enrichment score for each
chemical, returning a matrix (chemicals in rows, samples
in columns) suitable for clustering, association tests
against phenotypes, survival analysis, or heatmap
visualization.

```{r gsva}
gsva_scores <- enrichment_CTD(expr, method = "GSVA")
dim(gsva_scores)  # chemicals x samples
head(gsva_scores, 3)
```

Tune the underlying `GSVA::gsvaParam()` through `...`,
for example to restrict to gene sets of a given size:

```{r gsva-tune}
gsva_strict <- enrichment_CTD(
    expr, method = "GSVA",
    minSize = 5, maxSize = 500
)
nrow(gsva_strict)
```

## Visualizing results

`plot_CTD()` creates publication-ready plots from enrichment
results. It auto-detects the method: bar/dot plots of fold
enrichment for ORA/GSEA, bar/dot plots of `-log10(padj)`
coloured by direction for CAMERA, and a sample-level
heatmap of the top-variance chemicals for GSVA.

```{r plot-bar, fig.width=7, fig.height=5}
# ORA bar plot of top enriched chemicals
plot_CTD(ora_results, type = "bar")
```

```{r plot-dot, fig.width=7, fig.height=5}
# ORA dot plot (size = gene count, color = adjusted p-value)
plot_CTD(ora_results, type = "dot", n = 10)
```

```{r plot-camera, fig.width=7, fig.height=5}
# CAMERA bar plot: x-axis = -log10(padj), fill = Direction
plot_CTD(camera_results, type = "bar")
```

```{r plot-gsva, fig.width=7, fig.height=5}
# GSVA heatmap: top-variance chemicals across samples
plot_CTD(gsva_scores)
```

## Adjusting for multiple testing

By default, `enrichment_CTD()` uses the Benjamini-Hochberg
(`"BH"`) method for p-value adjustment. You can change this
via the `pAdjustMethod` parameter:

```{r padj-methods}
# Bonferroni correction (more conservative)
ora_bonf <- enrichment_CTD(genes, method = "ORA",
    pAdjustMethod = "bonferroni")

# No adjustment
ora_raw <- enrichment_CTD(genes, method = "ORA",
    pAdjustMethod = "none")

# Compare the adjusted p-value of the top hit
data.frame(
    method = c("BH (default)", "bonferroni", "none"),
    top_padj = c(min(ora_results$padj),
        min(ora_bonf$padj), min(ora_raw$padj))
)
```

Available methods: `"BH"` (default), `"bonferroni"`,
`"fdr"` (alias for BH), `"none"`.

# Choosing among the four methods

| Aspect | ORA | GSEA | CAMERA | GSVA |
|---|---|---|---|---|
| **Input** | Gene list | Ranked gene list | Expression matrix + design + contrast | Expression matrix |
| **Question** | Over-represented? | Cluster at extremes? | Stronger signal than rest of transcriptome (correlation-corrected)? | Per-sample chemical score |
| **Output** | Ranked chemicals (data.frame) | Ranked chemicals (data.frame) | Ranked chemicals (data.frame) | Chemical x sample score matrix |
| **Best for** | DEG lists | Exploratory, ranked | Multi-sample DE experiments, co-regulated gene sets | Patient stratification, downstream tests, heatmaps |

Use **ORA** when you have a well-defined gene list above a
significance threshold; **GSEA** to leverage the full
ranking without a cutoff; **CAMERA** when you have a
proper multi-sample design and want to control the
inflated false-positive rate that ORA/GSEA exhibit on
strongly co-regulated gene sets; **GSVA** when you want a
per-sample score to feed into downstream analyses
(clustering, survival, correlation with phenotypes).

# Updating the CTD data

CTD releases updated data periodically. To update:

1. Download the latest `CTD_chem_gene_ixns.csv.gz` from
    <https://ctdbase.org/reports/CTD_chem_gene_ixns.csv.gz>.
2. Decompress and re-run `import_CTD()` -- existing
    cache files are overwritten.

```{r reimport, eval=FALSE}
import_CTD("~/Downloads/CTD_chem_gene_ixns.csv")
```

# Session info

```{r session-info}
sessionInfo()
```
