---
title: "Introduction to ctdR"
author:
  - name: Luigi Corsaro
    affiliation: ""
    orcid: 0000-0003-1218-230X
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to ctdR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

**ctdR** identifies chemicals significantly associated with a set of genes
using data from the
[Comparative Toxicogenomics Database (CTD)](https://ctdbase.org).

Two enrichment methods are supported:

- **ORA** (Over-Representation Analysis) — tests whether the overlap between
  your gene list and each chemical's target genes is larger than expected by
  chance. Powered by
  [`clusterProfiler::enricher`](https://bioconductor.org/packages/clusterProfiler/).
- **GSEA** (Gene Set Enrichment Analysis) — detects chemicals whose targets
  cluster toward the extremes of a ranked gene list. Powered by
  [`fgsea::fgsea`](https://bioconductor.org/packages/fgsea/).

# Data licensing disclaimer

> **This package does NOT bundle, redistribute, or embed any data from the
> Comparative Toxicogenomics Database.** CTD data are created and maintained
> by NC State University and are subject to specific licensing terms. Users
> must download the data directly from <https://ctdbase.org> and comply with
> the [CTD Terms of Service](https://ctdbase.org/about/legal.jsp).

# Installation

## From GitHub

```{r install-github}
# install.packages("devtools")
devtools::install_github("drake69/ctdR")
```

## Bioconductor dependencies

ctdR depends on several Bioconductor packages. If they are not installed
automatically:

```{r install-bioc}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c(
  "fgsea", "org.Hs.eg.db", "clusterProfiler",
  "DOSE", "AnnotationDbi"
))
```

# Quick start

The ctdR workflow has three steps:

1. **Download** the CTD data file (once, manually)
2. **Import** the data into ctdR (once)
3. **Analyse** your gene list (as many times as needed)

## Step 1 — Download CTD data

Download **`CTD_chem_gene_ixns.csv.gz`** from:

<https://ctdbase.org/reports/CTD_chem_gene_ixns.csv.gz>

Then decompress it:
```bash
gunzip CTD_chem_gene_ixns.csv.gz
```

This produces `CTD_chem_gene_ixns.csv` (several GB uncompressed).

## Step 2 — Import into ctdR

```{r import}
library(ctdR)

import_CTD("~/Downloads/CTD_chem_gene_ixns.csv")
```

`import_CTD()` performs the following:

1. Reads the CSV (skipping the 27 CTD header lines).
2. Filters interactions to **Homo sapiens** only (OrganismID 9606).
3. Collects Entrez gene IDs for each chemical.
4. Maps Entrez IDs to HGNC gene symbols via `org.Hs.eg.db`.
5. Caches the processed data locally (under `rappdirs::user_cache_dir("ctdR")`).

This step takes several minutes depending on the CTD release size. You only
need to run it once — or again when you download a newer CTD release.

## Step 3 — Run enrichment analysis

### Prepare your gene list

Your input must be a data frame with at least two columns:

| Column         | Description                                                      |
|----------------|------------------------------------------------------------------|
| `entrez_ids`   | Character or numeric Entrez gene IDs                             |
| *(2nd column)* | Numeric value per gene (e.g. p-value, log fold-change)           |

The second column is used for ranking in GSEA and is ignored in ORA.

```{r prepare-genes}
genes <- data.frame(
  entrez_ids = c("7124", "3569", "7157", "672", "1956",
                 "4609", "3845", "207", "5290", "3553"),
  pvalue     = c(0.001, 0.003, 0.005, 0.008, 0.01,
                 0.02,  0.03,  0.04,  0.05,  0.06)
)
```

### Over-Representation Analysis (ORA)

ORA tests whether the overlap between your gene list and each chemical's
known gene targets is significantly larger than expected by chance.

```{r ora}
ora_results <- enrichment_CTD(genes, method = "ORA")
head(ora_results)
```

**ORA output columns:**

| Column           | Description                                    |
|------------------|------------------------------------------------|
| `ChemicalID`     | CTD chemical identifier (e.g. `"D000082"`)     |
| `ChemicalName`   | Human-readable chemical name                   |
| `GeneRatio`      | Proportion of input genes in the chemical's set|
| `BgRatio`        | Background ratio                               |
| `pvalue`         | Raw p-value                                    |
| `padj`           | Adjusted p-value (BH method)                   |
| `foldEnrichment` | GeneRatio / BgRatio                            |
| `geneID`         | Enriched gene symbols                          |
| `Count`          | Number of overlapping genes                    |

### Gene Set Enrichment Analysis (GSEA)

GSEA uses the full ranked gene list (ranked by the numeric column) to detect
chemicals whose targets cluster toward the top or bottom of the ranking.

```{r gsea}
gsea_results <- enrichment_CTD(genes, method = "GSEA")
head(gsea_results)
```

**GSEA output columns:**

| Column           | Description                                |
|------------------|--------------------------------------------|
| `ChemicalID`     | CTD chemical identifier                    |
| `ChemicalName`   | Human-readable chemical name               |
| `pval`           | Raw p-value                                |
| `padj`           | Adjusted p-value                           |
| `ES`             | Enrichment score                           |
| `NES`            | Normalized enrichment score                |
| `size`           | Size of the gene set                       |
| `leadingEdge`    | Leading-edge gene subset                   |
| `foldEnrichment` | |ES| / mean(ES)                           |
| `Enriched_GENE`  | Comma-separated enriched gene symbols      |

# Choosing between ORA and GSEA

| Aspect        | ORA                              | GSEA                              |
|---------------|----------------------------------|-----------------------------------|
| **Input**     | Gene list (no ranking needed)    | Ranked gene list                  |
| **Question**  | Are these genes over-represented?| Do targets cluster at the extremes?|
| **Threshold** | Requires a significance cutoff   | Uses full ranking, no cutoff      |
| **Best for**  | DEG lists with clear cutoff      | Exploratory, ranked data          |

Use **ORA** when you have a well-defined gene list (e.g. differentially
expressed genes above a significance threshold). Use **GSEA** when you want
to leverage the full ranking without imposing an arbitrary threshold.

# Updating the CTD data

CTD releases updated data periodically. To update:

1. Download the latest `CTD_chem_gene_ixns.csv.gz` from
   <https://ctdbase.org/reports/CTD_chem_gene_ixns.csv.gz>.
2. Decompress and re-run `import_CTD()` — existing cache files are overwritten.

```{r reimport}
import_CTD("~/Downloads/CTD_chem_gene_ixns.csv")
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
