---
title: "From RNA-seq to chemical enrichment: a complete workflow"
author:
  - name: Luigi Corsaro
    orcid: 0000-0003-1218-230X
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>",
    message  = FALSE,
    warning  = FALSE
)
```

# Biological context

Dexamethasone is a synthetic glucocorticoid widely used as an
anti-inflammatory and immunosuppressive agent. Its transcriptomic
effects on immune cells are well characterised: it suppresses
pro-inflammatory cytokines, reprograms monocyte and lymphocyte gene
programmes, and leaves a distinctive signature in peripheral blood
mononuclear cells (PBMCs).

This tutorial asks a straightforward question: **given the RNA-seq
signature of dexamethasone treatment in human PBMCs, which chemicals in
the Comparative Toxicogenomics Database (CTD) share that signature?**

We answer it using all four methods provided by ctdR — GSEA, ORA,
CAMERA, and GSVA — and show how to visualise and interpret each result.

**Data source**: we use
[GSE311566](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE311566),
a public GEO series comparing the effects of PFAS chemicals and
dexamethasone on human PBMCs from female donors. We work with the
Dexamethasone vs. vehicle (DMSO/Ctrl) contrast, female samples only (7
samples total: 4 Ctrl + 3 Dex).

------------------------------------------------------------------------

> **What this tutorial covers**
>
> - Downloading and preparing public RNA-seq count data from GEO
> - Differential expression with `limma`
> - Chemical enrichment analysis (GSEA, ORA, CAMERA, GSVA) using `ctdR`
> - Visualisation with `plot_CTD()`
>
> **What this tutorial does NOT cover**
>
> - Raw read quality control (FastQC, trimming)
> - Alignment and count generation
> - Normalisation strategy selection (TMM, VST, etc.)
> - Batch correction or covariate adjustment
> - Biological replicate adequacy or power analysis
>
> This is an end-to-end illustration of the `ctdR` workflow on publicly
> available processed data, not a template for a production differential
> expression pipeline.

------------------------------------------------------------------------

# Setup

```{r libraries}
library(ctdR)
library(limma)
library(org.Hs.eg.db)
library(AnnotationDbi)
```

# Step 1 — Download data from GEO

The normalised count matrix for GSE311566 is available directly from
NCBI FTP. We download it once and cache it in the session temporary
directory; re-running this chunk skips the download if the file already
exists.

```{r download}
GEO_URL <- paste0(
    "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE311nnn/GSE311566/",
    "suppl/GSE311566_PBMCs_Female_normalized_counts.txt.gz"
)
tmp_gz <- file.path(tempdir(), "GSE311566_Female_normalized_counts.txt.gz")
if (!file.exists(tmp_gz)) {
    message("Downloading GSE311566 from NCBI FTP ...")
    utils::download.file(GEO_URL, tmp_gz, mode = "wb", quiet = FALSE)
}
```

# Step 2 — Prepare the expression matrix

```{r prepare-counts}
counts <- utils::read.table(
    gzfile(tmp_gz),
    header = TRUE, sep = "\t", row.names = 1,
    check.names = FALSE, stringsAsFactors = FALSE
)
counts$DESCRIPTION <- NULL  # all NA in this dataset

# Female Ctrl and DEX samples only
keep_cols <- grep("^(Ctrl|DEX)_F_", colnames(counts), value = TRUE)
counts <- counts[, keep_cols]
group <- factor(
    ifelse(grepl("^DEX_", keep_cols), "Dex", "DMSO"),
    levels = c("DMSO", "Dex")
)
message(nrow(counts), " genes, ", ncol(counts), " samples: ",
    paste(table(group), names(table(group)), sep = " ", collapse = " / "))
```

The GEO file uses Ensembl gene IDs. We strip version suffixes and map to
Entrez IDs, which are the identifier type required by `ctdR`.

```{r map-entrez}
ensg <- sub("\\..*$", "", rownames(counts))
entrez_map <- AnnotationDbi::mapIds(
    org.Hs.eg.db,
    keys     = ensg,
    keytype  = "ENSEMBL",
    column   = "ENTREZID",
    multiVals = "first"
)
keep_g <- !is.na(entrez_map) & !duplicated(entrez_map)
counts <- counts[keep_g, ]
rownames(counts) <- entrez_map[keep_g]
expr <- log2(as.matrix(counts) + 1)
message(nrow(expr), " genes retained after Entrez mapping")
```

# Step 3 — Import CTD data

```{r import-ctd}
ctd_csv <- "~/Downloads/CTD_chem_gene_ixns.csv"
ctd_gz  <- "~/Downloads/CTD_chem_gene_ixns.csv.gz"

if (file.exists(ctd_csv)) {
    message("Using full CTD: ", ctd_csv)
    import_CTD(ctd_csv)
} else if (file.exists(ctd_gz)) {
    message("Using full CTD (compressed): ", ctd_gz)
    import_CTD(ctd_gz)
} else {
    message(
        "Full CTD not found in ~/Downloads/ — using bundled toy sample ",
        "(10 chemicals). Results will be illustrative only.\n",
        "Download CTD_chem_gene_ixns.csv.gz from ctdbase.org for a ",
        "powered analysis."
    )
    import_CTD(system.file(
        "extdata", "CTD_chem_gene_ixns_sample.csv",
        package = "ctdR"
    ))
}
```

> **Note on CTD coverage**: this tutorial uses the small toy dataset
> bundled with `ctdR` (10 chemicals, 17 genes). It is sufficient to
> demonstrate the API and visualisations. For a powered analysis capable
> of discovering novel chemical associations, download the full
> `CTD_chem_gene_ixns.csv.gz` from
> <https://ctdbase.org/reports/CTD_chem_gene_ixns.csv.gz> and run
> `import_CTD()` on that file instead.

# Step 4 — Differential expression with limma

We fit a linear model and apply empirical Bayes moderation. The design
matrix has two columns: the intercept (DMSO baseline) and `groupDex`
(Dex vs DMSO contrast, coefficient 2).

```{r limma-de}
design <- model.matrix(~ group)
colnames(design)

fit <- limma::lmFit(expr, design)
fit <- limma::eBayes(fit)

de <- limma::topTable(fit, coef = 2, n = Inf, sort.by = "P")
de$EntrezID <- rownames(de)

head(de[, c("EntrezID", "logFC", "t", "P.Value", "adj.P.Val")])
```

A brief look at significance:

```{r de-summary}
c(
    total_genes   = nrow(de),
    padj_lt_05    = sum(de$adj.P.Val < 0.05),
    pvalue_lt_05  = sum(de$P.Value  < 0.05)
)
```

# Step 5 — Chemical enrichment

## GSEA — Gene Set Enrichment Analysis

GSEA uses the **full ranked gene list** (here ranked by nominal p-value;
internally converted to `-log10(p)` as the running-sum statistic). It
detects chemicals whose known target genes collectively cluster toward
the most significant end of the ranking, without requiring a
significance threshold.

```{r gsea}
gsea_input <- data.frame(
    EntrezID = de$EntrezID,
    pvalue   = de$P.Value,
    stat     = de$t,
    row.names = NULL
)
gsea_results <- enrichment_CTD(gsea_input, method = "GSEA")
head(gsea_results[, c(
    "ChemicalName", "PValue", "PValueAdjusted",
    "NormalizedEnrichmentScore", "FoldEnrichment"
)])
```

`FoldEnrichment` here is `|ES| / mean(|ES|)` across all tested chemicals
— a relative measure of how strongly enriched this chemical's gene set
is compared to the average. `NES` (Normalized Enrichment Score) accounts
for gene set size differences and is the primary statistic for ranking
chemicals.

```{r plot-gsea, fig.width=7, fig.height=5}
plot_CTD(gsea_results, type = "bar", n = 10,
         title = "GSEA — top chemicals by NES")
```

```{r plot-gsea-dot, fig.width=7, fig.height=5}
plot_CTD(gsea_results, type = "dot", n = 10,
         title = "GSEA — top chemicals (dot)")
```

### Where is Dexamethasone?

The top-ranked chemical may or may not be Dexamethasone depending on
which CTD dataset is loaded. With the full CTD (\~11 000 chemicals and
strict BH correction) Dexamethasone competes against many other
chemicals and may rank lower than expected, or may not reach
significance after correction. The code below locates it regardless of
its rank:

```{r gsea-dex}
dex_idx  <- grep("dexamethasone", gsea_results$ChemicalName,
                 ignore.case = TRUE)
if (length(dex_idx) > 0) {
    cat("Dexamethasone: rank", dex_idx, "of", nrow(gsea_results), "\n")
    gsea_results[dex_idx, c("ChemicalName", "PValue", "PValueAdjusted",
                             "NormalizedEnrichmentScore", "FoldEnrichment")]
} else {
    message("Dexamethasone not found — it may be absent from the loaded CTD cache.")
}
```

> **Why Dexamethasone may not top the list with the full CTD:**
>
> 1.  **Multiple testing burden** — 11 000 tests vs 10 in the toy
>     sample; the BH threshold is \~1 000× more stringent.
> 2.  **Larger gene sets** — in the full CTD, Dexamethasone has hundreds
>     of annotated targets; very large sets dilute the GSEA signal in
>     small cohorts (7 samples here).
> 3.  **Competing chemicals** — other chemicals may have gene sets that
>     overlap more tightly with the top-ranked DE genes.

## ORA — Over-Representation Analysis

ORA works on a **binary gene list**: genes declared significant vs the
rest of the universe. It asks whether the overlap between your
significant genes and each chemical's known targets is larger than
expected by chance (hypergeometric test).

```{r ora}
sig <- de[de$adj.P.Val < 0.05, c("EntrezID", "P.Value")]
colnames(sig)[2] <- "pvalue"
message(nrow(sig), " genes at adj.P.Val < 0.05")

ora_results <- enrichment_CTD(sig, method = "ORA")
```

```{r ora-results}
if (nrow(ora_results) > 0) {
    head(ora_results[, c(
        "ChemicalName", "PValue", "PValueAdjusted",
        "FoldEnrichment", "Count", "EnrichedGenes"
    )])
} else {
    message(
        "ORA returned 0 results — expected with the toy CTD sample.\n",
        "Run import_CTD() on the full CTD file for a powered analysis."
    )
}
```

```{r ora-dex}
if (nrow(ora_results) > 0) {
    dex_idx <- grep("dexamethasone", ora_results$ChemicalName,
                    ignore.case = TRUE)
    if (length(dex_idx) > 0) {
        cat("Dexamethasone: rank", dex_idx, "of", nrow(ora_results), "\n")
        ora_results[dex_idx, c("ChemicalName", "PValue", "PValueAdjusted",
                                "FoldEnrichment", "Count")]
    } else {
        message("Dexamethasone not found in ORA results.")
    }
}
```

```{r plot-ora, eval=nrow(ora_results) > 0, fig.width=7, fig.height=5}
plot_CTD(ora_results, type = "dot", n = 10,
         title = "ORA — top chemicals by fold enrichment")
```

`FoldEnrichment` in ORA is `GeneRatio / BackgroundRatio`: the proportion
of your significant genes that are targets of this chemical, divided by
the same proportion in the background universe. A value greater than 1
indicates over-representation.

## CAMERA — Competitive gene-set test

CAMERA tests, for each chemical, whether its target genes show a
**stronger differential signal** than the rest of the transcriptome,
while correcting for inter-gene correlation within the gene set. It is
particularly well suited to expression matrix experiments with a clear
design and contrast.

We reuse the `design` matrix built for limma. `contrast = 2` selects the
`groupDex` coefficient (Dex vs DMSO).

```{r camera}
camera_results <- enrichment_CTD(
    expr,
    method   = "CAMERA",
    design   = design,
    contrast = 2
)
head(camera_results[, c(
    "ChemicalName", "PValue", "PValueAdjusted",
    "GeneSetSize", "Direction"
)])
```

Unlike ORA and GSEA, CAMERA does not report a `FoldEnrichment` value —
it reports `Direction` (`"Up"` or `"Down"`), indicating whether the
chemical's target genes are predominantly up- or down-regulated under
the tested contrast.

```{r plot-camera, fig.width=7, fig.height=5}
plot_CTD(camera_results, type = "bar", n = 10,
         title = "CAMERA — direction-aware chemical enrichment")
```

```{r camera-dex}
dex_idx <- grep("dexamethasone", camera_results$ChemicalName,
                ignore.case = TRUE)
if (length(dex_idx) > 0) {
    cat("Dexamethasone: rank", dex_idx, "of", nrow(camera_results), "\n")
    camera_results[dex_idx, c("ChemicalName", "PValue", "PValueAdjusted",
                               "GeneSetSize", "Direction")]
} else {
    message("Dexamethasone not found in CAMERA results.")
}
```

## GSVA — Gene Set Variation Analysis

GSVA produces a **per-sample enrichment score** for each chemical: a
matrix with chemicals in rows and samples in columns. There is no single
group-level p-value; instead, these scores can be used for clustering,
heatmap visualisation, survival association, or as input to downstream
tests.

> **Computation time**: GSVA scores all chemicals across all samples. On
> the full expression matrix this may take a few minutes. Use
> `BiocParallel::register()` to speed it up on multi-core machines.

```{r gsva}
gsva_scores <- enrichment_CTD(expr, method = "GSVA")
dim(gsva_scores)  # chemicals x samples
```

```{r plot-gsva, fig.width=7, fig.height=6}
plot_CTD(gsva_scores,
         title = "GSVA — per-sample chemical enrichment scores")
```

The heatmap rows are chemicals ranked by across-sample variance
(top-variance chemicals shown first). Samples that cluster together
share a similar chemical-exposure transcriptomic profile.

To inspect Dexamethasone's per-sample scores directly:

```{r gsva-dex}
dex_idx <- grep("dexamethasone", rownames(gsva_scores),
                ignore.case = TRUE)
if (length(dex_idx) > 0) {
    cat("Dexamethasone row(s):", rownames(gsva_scores)[dex_idx], "\n")
    gsva_scores[dex_idx, , drop = FALSE]
} else {
    message("Dexamethasone not found in GSVA score matrix.")
}
```

Positive scores indicate that Dexamethasone's target genes are
collectively up-regulated in that sample relative to the background;
negative scores indicate down-regulation. A clear separation between Dex
and Ctrl samples (positive scores in Dex, negative in Ctrl, or vice
versa) would support the expected glucocorticoid signature.

# Step 6 — Comparing results across methods

Each method answers a subtly different question:

| Method | Input | Question answered | Key output column |
|------------------|------------------|------------------|------------------|
| GSEA | Full ranked list | Do this chemical's targets cluster at the extreme of the ranking? | `NES`, `FoldEnrichment` |
| ORA | Significant gene list | Is the overlap larger than expected by chance? | `FoldEnrichment`, `Count` |
| CAMERA | Expression matrix + design | Is the differential signal stronger than the transcriptome average? | `Direction` |
| GSVA | Expression matrix | What is the per-sample enrichment score? | score matrix |

GSEA and ORA both report `FoldEnrichment`, but it means different
things: in GSEA it is relative to the mean ES across chemicals
(`|ES| / mean(|ES|)`); in ORA it is the ratio of observed to expected
overlap (`GeneRatio / BackgroundRatio`). Do not compare these numbers
across methods.

For a discovery workflow on a single contrast, **GSEA is recommended as
the primary method** — it uses all genes and is robust to the choice of
significance threshold. ORA is useful as a complementary check when a
reliable binary gene list is available. CAMERA is the method of choice
when inter-gene correlation is a concern. GSVA is best reserved for
multi-sample stratification or when you want to visualise chemical
signatures across a cohort.

# Dexamethasone — ranking recap across all methods

Since Dexamethasone is the treatment used in GSE311566, we expect its
CTD gene set to carry a strong signal in this dataset. The code below
collects its position and key statistics from each method into a single
summary, regardless of whether it reached the top rank.

```{r dex-recap-gsea}
dex_gsea <- local({
    idx <- which(tolower(gsea_results$ChemicalName) == "dexamethasone")
    if (length(idx) == 0L) return(NULL)
    r <- gsea_results[idx, ]
    data.frame(
        Method    = "GSEA",
        Rank      = idx,
        Total     = nrow(gsea_results),
        PValue    = signif(r$PValue, 3),
        PAdj      = signif(r$PValueAdjusted, 3),
        StatLabel = "NES",
        Stat      = signif(r$NormalizedEnrichmentScore, 3)
    )
})
```

```{r dex-recap-ora}
dex_ora <- local({
    if (nrow(ora_results) == 0L) return(NULL)
    idx <- which(tolower(ora_results$ChemicalName) == "dexamethasone")
    if (length(idx) == 0L) return(NULL)
    r <- ora_results[idx, ]
    data.frame(
        Method    = "ORA",
        Rank      = idx,
        Total     = nrow(ora_results),
        PValue    = signif(r$PValue, 3),
        PAdj      = signif(r$PValueAdjusted, 3),
        StatLabel = "FoldEnrichment",
        Stat      = signif(r$FoldEnrichment, 3)
    )
})
```

```{r dex-recap-camera}
dex_camera <- local({
    idx <- which(tolower(camera_results$ChemicalName) == "dexamethasone")
    if (length(idx) == 0L) return(NULL)
    r <- camera_results[idx, ]
    data.frame(
        Method    = "CAMERA",
        Rank      = idx,
        Total     = nrow(camera_results),
        PValue    = signif(r$PValue, 3),
        PAdj      = signif(r$PValueAdjusted, 3),
        StatLabel = "Direction",
        Stat      = r$Direction
    )
})
```

```{r dex-recap-gsva}
dex_gsva <- local({
    idx <- which(tolower(rownames(gsva_scores)) == "dexamethasone")
    if (length(idx) == 0L) return(NULL)
    scores       <- gsva_scores[idx, , drop = FALSE]
    dex_samples  <- grepl("^DEX_",  colnames(gsva_scores))
    ctrl_samples <- grepl("^Ctrl_", colnames(gsva_scores))
    mean_dex     <- mean(scores[, dex_samples])
    mean_ctrl    <- mean(scores[, ctrl_samples])
    data.frame(
        Method    = "GSVA",
        Rank      = NA_integer_,
        Total     = nrow(gsva_scores),
        PValue    = NA_real_,
        PAdj      = NA_real_,
        StatLabel = "mean(Dex) - mean(Ctrl)",
        Stat      = signif(mean_dex - mean_ctrl, 3)
    )
})
```

```{r dex-recap-table}
recap <- do.call(rbind, Filter(Negate(is.null),
                               list(dex_gsea, dex_ora, dex_camera, dex_gsva)))

if (is.null(recap) || nrow(recap) == 0L) {
    message("Dexamethasone not found in any method's results.")
} else {
    recap$RankOf <- ifelse(
        is.na(recap$Rank), "—",
        paste0(recap$Rank, " / ", recap$Total)
    )
    print(recap[, c("Method", "RankOf", "PValue", "PAdj",
                    "StatLabel", "Stat")],
          row.names = FALSE)
}
```

`RankOf` is the position of Dexamethasone in each result table sorted
by adjusted p-value (GSEA, ORA, CAMERA) or — for GSVA — the
difference between the mean per-sample score in treated vs control
samples (positive = target genes up-regulated in Dex samples).

# Session information

```{r sessionInfo}
sessionInfo()
```
